---
title: "Movielens"
date: '`r Sys.time()`'
output:
  html_document:
    #code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
library(scales)
library(tidyverse)
library(knitr)

# set plot theme
theme_set(theme_bw())
```

# Load and preview data

Read data from the `ratings.csv` file
```{r load-data}
ratings <- read_csv("week3/ratings.csv",
  col_names = c("user_id", "movie_id", "rating", "timestamp")
)
head(ratings)
```

Loaded `r format(object.size(ratings), units="Mb")` of ratings data, containing `r format(nrow(ratings), big.mark = ",")` ratings. Here's a preview:
```{r preview-data}
head(ratings) %>% kable()
```

# Summary statistics

```{r dist-ratings}
# plot the distribution of rating values https://speakerdeck.com/jhofman/modeling-social-data-lecture-2-introduction-to-counting?slide=26

rating_values <- ggplot(ratings, aes(rating)) +
  geom_bar() +
  labs(
    x = "Ratings",
    y = "No of Ratings"
  )
rating_values
```

## Per-movie stats

```{r aggregate-by-movie}
# aggregate ratings by movie, computing mean rating and number of ratings
# hint: use the n() function for easy counting within a group
aggregate <- ratings %>%
  group_by(movie_id) %>%
  summarise(
    mean_rating = mean(rating),
    no_of_ratings = n()
  )

aggregate
```

```{r dist-movie-popularity}
# plot distribution of movie popularity (= number of ratings the movie received)
# hint: try scale_x_log10() for a logarithmic x axis

ggplot(aggregate, aes(no_of_ratings)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Distribution of Movie Popularity",
    x = "No of ratings",
    y = "No of movies"
  )

```

```{r dist-mean-ratings-by-movie}
# plot distribution of mean ratings by movie https://speakerdeck.com/jhofman/modeling-social-data-lecture-2-introduction-to-counting?slide=28
# hint: try geom_histogram and geom_density
ggplot(aggregate, aes(mean_rating)) +
  geom_density(fill = "black")

ggplot(aggregate, aes(mean_rating)) +
  geom_histogram() +
  geom_density(fill = "black")
```

```{r cdf-movie-pop}
# rank movies by popularity (number of ratings) and compute the cdf, or fraction of all views covered by the top-k movies https://speakerdeck.com/jhofman/modeling-social-data-lecture-2-introduction-to-counting?slide=30
# hint: use dplyr's rank and arrange functions, and the base R sum and cumsum functions
# store the result in a new data frame so you can use it in creating figure 2 from the paper below

cdf_df <- aggregate %>%
  arrange(desc(no_of_ratings)) %>%
  mutate(
    rank = row_number(),
    cum_ratings = cumsum(no_of_ratings),
    cdf = cum_ratings / sum(no_of_ratings)
  )
cdf_df

  ggplot(cdf_df, aes(x = rank, y = cdf)) +
    geom_line() +
    labs(
      title = "CDF of Movie Popularity",
      x = "Top-k Movies (Ranked by Popularity)",
      y = "Cumulative Fraction of All Ratings"
    ) +
    theme_minimal()

```


# Per-user stats

```{r aggregate-by-user}
# aggregate ratings by user, computing mean and number of ratings
per_user_stats <- ratings %>%
  group_by(user_id) %>%
  summarise(
    mean_rating = mean(rating),
    no_of_ratings = n()
  )
per_user_stats

```

```{r dist-user-activity}
# plot distribution of user activity (= number of ratings the user made)
# hint: try a log scale here
ggplot(per_user_stats, aes(x = no_of_ratings)) +
  geom_histogram(color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of User Activity",
    x = "Number of Ratings (Log Scale)",
    y = "Number of Users"
  )
```

# Anatomy of the long tail

```{r long-tail}
# generate the equivalent of figure 2a of this paper:
# note: don't worry about the "null model" lines
# just do the solid lines and dotted line (optional)
# https://5harad.com/papers/long_tail.pdf

# Specifically, for the subset of users who rated at least 10 movies,
# produce a plot that shows the fraction of users satisfied (vertical
# axis) as a function of inventory size (horizontal axis). We will
# define "satisfied" as follows: an individual user is satisfied p% of
# the time at inventory of size k if at least p% of the movies they
# rated are contained in the top k most popular movies. As in the
# paper, produce one curve for the 100% user satisfaction level and
# another for 90%---do not, however, bother implementing the null
# model (shown in the dashed lines).

```{r long-tail, message=FALSE, warning=FALSE}

ranked_ratings <- ratings %>%
  left_join(movie_popularity, by = "movie_id")

ranked_ratings

user_movie_ranks <- ranked_ratings %>%
  group_by(user_id) %>%
  summarise(ranks = list(rank), .groups = "drop")
user_movie_ranks


# Compute each user's satisfaction thresholds
user_satisfaction <- user_movie_ranks %>%
  mutate(
    threshold_100 = sapply(ranks, max),
    threshold_90 = sapply(ranks, function(x) quantile(x, probs = 0.9))
  )

# Define inventory sizes from 1 to max needed
inventory_sizes <- 1:max(user_satisfaction$threshold_100)

# Compute 100% and 90% satisfaction at each inventory size
satisfaction_100 <- sapply(inventory_sizes, function(k) {
  mean(user_satisfaction$threshold_100 <= k)
})

satisfaction_90 <- sapply(inventory_sizes, function(k) {
  mean(user_satisfaction$threshold_90 <= k)
})

# Create data frame for plotting
satisfaction_df <- tibble(
  inventory_size = inventory_sizes,
  satisfaction_100 = satisfaction_100,
  satisfaction_90 = satisfaction_90
)

# Plot
ggplot(satisfaction_df, aes(x = inventory_size)) +
  geom_line(aes(y = satisfaction_100)) +
  geom_line(aes(y = satisfaction_90)) +
  labs(
    title = "User Satisfaction vs Inventory Size",
    x = "Inventory Size (Top-k Popular Movies)",
    y = "Fraction of Users Satisfied"
  ) +
  theme_minimal()



```
